Towards an Ecological Trait-data Standard - Appendix A

### 2018-05-18 manuscript/appendix_A.Rmd
add Appendix A: Rmarkdown file transfer of google spreadsheet to pdf

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2] \footnotemark[3]
- name: Malte Jochum \footnotemark[4]
- name: Gaetane Le Provost \footnotemark[3]
- name: Andreas Ostrowski \footnotemark[5]
- name: Caterina Penone \footnotemark[4]
- name: David Fichtmüller \footnotemark[5]
- name: Anton Güntsch \footnotemark[5]
- name: Pete Manning \footnotemark[3]
- name: Birgitta König-Ries\footnotemark[6]
- name: Nadja K. Simons \footnotemark[7]
csl: amnat.csl
fontsize: 12pt
geometry: margin=1.3in
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

\linenumbers

# Abstract

1.	Trait-based approaches are widespread throughout ecological research, due to the great potential for trait data to deliver general and mechanistic conclusions. Accordingly, a wealth of trait data is available for many organism groups but, due to a lack of standardisation, these data come in heterogeneous formats.
2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories.
3. In order to facilitate the standardisation and harmonisation of distributed trait datasets, we propose a general and simple vocabulary as well as a simple data structure for storing and sharing ecological trait data. 
4. Additionally, we provide an R-package that enables the transformation of any dataset into the proposed format. This also allows trait datasets from heterogeneous sources to be harmonized and merged, thus facilitating data compilation for any particular research focus. 
5. With these decentralised tools for trait-data harmonisation, we intend to facilitate the exchange and analysis of trait data within ecological research and enable global syntheses of traits across a wide range of taxa and ecosystems.

# Key-words:

functional ecology, species traits, semantic web, ontologies, data standard

# Introduction

<!-- CP:  I think we should develop more on what are the current problems/limitations (quickly because it’s then expanded in the paper) so it becomes evident that the standard+package fill a real need.

That’s the structure I try to follow:
1. Traits are important in ecology, evolution and conservation (i.e.  It is important to mention that traits are important in research and give some examples of questions that traits can help to solve but maybe not spend too long on it (most people know that they are important).

2. To increase the scale of studies (spatial, temporal, taxonomical) we need to put together trait data from different sources. When we do this we can answer new questions and find generalities. So there is a growing interest in assembling trait data from different sources and taxa (e.g. give the number of papers using TRY).

3. There is a problem of standardization that makes it difficult to match trait datasets from different sources. Different reasons: no traits definitions, no common structure.

4. One solution are databases. But these are costly and depend on the will of few persons/funding. They also cover few taxonomic groups. Matching between different databases is still difficult because of different definitions and structures. Most data are still uploaded separately.
We need a solution to make their use easy. 
-->
<!-- GLP: 
 Add a general paragraph on the fact that there have been some initiatives for the development of trait data base for plants (e.g. TRY), why is it important (provide quick access to plant trait values, allowing comparisons even between ecosystems and biomes), now the development of trait data base for other organisms (from microbes to mammals). Say also something on the development of general protocols for measuring traits (e.g. Moretti et al. 2016; Freschet et al. 2017) ?

Importance of data standardization. This is the main idea here and so I wonder if some idea of the first point (Need for data standardization) should not be merge here with the last paragraph of the introduction ? As there is also some repetitions between those 2 parts.

I wonder if you should speak about ‘Functional traits’here or just Traits ? As you are not only considered Functional traits in this paper (e.g. life history traits) ? Maybe remove the definition no ? 
-->
<!-- PM: Thuis is quite complex- e.g. ‘taxonomic impediment’etc. I actually think you can simplify this and built these references into the more general text above by adding a copule of more examples. I don’t think you need to say much about the merits of the trait based approach here- it’s covered elsewhere.

Huge number of references here- I think you can strip it down to a few key ones, and reviews. Overall I think this section is quite a lot more complex than it needs to be – I think all you need to say is that it’s very popular, general and mechanistic and so adopted throughout eco and evo, with no sign of slowing down- the TRY database download stats can provide a quantitative measure of this- see their site. 
-->

<!-- 1. Traits are important in ecology, evolution and conservation -->

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. Because trait-based approaches to ecological and evolutionary research questions allow studying both patterns and mechanisms [@lavorel02; @diaz16], recent years have seen a proliferation of trait-based research in a wide range of fields.
Trait-based studies have been conducted in a wide range of thematic areas ranging from the evolutionary basis of individual-level properties [@salguero-gomez16] to global patterns of biodiversity [@diaz16] and ecosystem functioning [@allan15; @bello10]. With this mechanistic depth, the trait framework relates losses of ecosystem function to changes in biodiversity, and on to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15]. <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
Using traits is also a promising means of bypassing taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@leprovost17; @duarte11; @schrodt15].

<!-- Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. 
Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [ @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13]. These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16]. <!-- GLP: Maybe not the best example as they are not looking at trait diversty and species interactions ? 
Put instead:
Gravel et al. 2016 The meaning of functional trait composition of food webs for ecosystem functioning
And Coux et al. 2016 Linking species functional roles to their network roles
?
-->
<!--
Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @simons16; @gossner16; @moretti13;  @kuhsel15;  @birkhofer14], as well as macro-patterns of biodiversity at the global scale [@diaz16; e.g. @reich97]. <!-- Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].--> <!-- GLP: What do you mean by processes ? I’m not sure that the literature cited is always appropriate here ?
-->
<!--
Most importantly, the trait framework relates key ecosystem services for human well-being [@diaz11; @lavorel12; @allan15] back to changes in biodiversity in consequence to global change or local anthropogenic land use [ @diaz07; @gossner16; @mangels17; @jennings09; @simons16;  @mouillot13;  @vandewalle10;  @birkhofer15;@gamez-virues15; @birkhofer17]. 
--> 
<!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, which continues to integrate a rapidly increasing number of fundamental and applied research questions. 
Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. -->

<!-- 2. assemble trait data into databases -->

Many issues in trait-based research arise when compiling datasets from several sources. Data may differ in taxonmic nomenclature and resolution (e.g. reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the methodology applied in measurements. All of these factors render trait data extremely heterogeneous and make the task of data compilation time-consuming or even prohibitive.
However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to acheive sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time.

<!-- 3. trait databases have been developed, with standard structure and terms --> 
To this end, trait data have been standardised and compiled in centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g. PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17]. 
These databases are then published as static data tables (i.e. product of a one-time effort as it is the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods (e.g. TRY, www.try-db.org). Furthermore, these portals map heterogeneous data into a common scheme, and importantly, also offer access control and data usage policies. As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers. 
Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed. These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17] for trait-data are beeing developed.
<!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->

<!-- 4. distributed trait data must be integrated --> 
While these community efforts for centralisation and harmonization are great advancements in biodiversity informatics and data-management, at the same time the number of decentralised trait-datasets is increasing rapidly. This is due to the fact that open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12; @swan12; @allison15]., which supports. As a result, an increasing number of individual research projects comply with open-data standards and publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material. <!-- The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.-->
It is likely that data will become increasingly available, but a lack of data standardisation will hamper the efficient re-use and synthesis of published datasets.

In this paper, we review existing trait databases and online portals, as well as initiatives for standardisation. We discuss current practice and the importance of data standards for trait-based research, and we identify current deficits in standardisation from a pragmatic view of data providers and data users. Based on these considerations, we propose a minimal structure and vocabulary for describing trait-datasets, that is compatible with existing terminology standards for biodiversity data. Finally, we present an R package that assists the harmonisation of trait data from distributed sources.

With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of trait-data standardisation and sketch a roadmap towards an accessible ecological trait data standard.

<!-- FDS: Why 'ecological' trait data? what is special about this subject? Environmental context, community context, space and time, attributes of functional role in the ecosystem and energetic demands, interactions (not covered right now?), correlation and economic trade-offs in trait evolution -->

<!-- PM: Can I suggest a glossary of terms – this would be really helpful for people who are not data management specialists

FDS: see Garnier et al. but may contain:

-->
\scriptsize

Term | Definition 
----- | ---------
Term | a word that describes a particular *concept* as part of the specialised vocabulary of a field
Concept | An idea, notion or object that is made explicit in a communication context by *name*, definition, *URI* or other reference (https://www.w3.org/TR/skos-reference/)
Controlled vocabulary | 
Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g. trait data), that is recognized by a large proportion of members of the application context
Terminology | The  body of *terms* used with a particular application in a subject of study, usually formalized in a *thesaurus* or *ontology*
Thesaurus | Controlled vocabulary that provides key *terms* with their associated *concepts* for a specific field or domain of interest [@laporte12] 
Ontology | Controlled vocabulary that (opposed to a *thesaurus*) relates *concepts* to each other by cross-references, e.g. defines a hierarchy of terms; thus a formal model of a the objects and their relationships in a domain of interest [@gruber95]
Semantic web | The concept of connecting data on the internet via well-defined standards and exchange protocols [@berners-lee02]; the World Wide Web Consortium (W3C) defines standards, i.e. specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/)
Dataset | a set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being harmonized across all data entries;  
Data table | a two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time
Database | a suite of data compiled from multiple *datasets*, i.e. from multiple study contexts or observation types; may take the form of a two-dimensional data table, but mostly is organised in into *relational databases* using database software tools; 
Relational database | Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
Online database | A relational database that is made accessible on the internet offering forms for filtering and downloading subsets of the data; often using technologies like SQL; some online databases offer access via a *webservice* and an *API* that can be addressed computationally 
Identifier (ID) | A unique label that relates entries within and across datasets; is used to link data tables into a relational database; can be user-specific or as a *URI* be part of a globally valid ontology or thesaurus  
Uniform Resource Identifier (URI) |
Web service | An exchange protocoll to access online databases directly and programmatically, i.e. by calls from a software tool; 
Application Programming Interface (API) | 
Online Portal | A website designed as a platform for the exchange of information, e.g. trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access
Repository | Archive for data files; online repositories make data available for public access, provide metadata and (not always) facilitate citations via DOIs (Digital Object Identifiers)
Metadata | Data documentation representing the higher level information or instructions that describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97]
Darwin Core Standard (Dwc) | Body  of standards containing a glossary of terms intended to facilitate the sharing of information about biological diversity by providing reference definitions, examples, and commentaries. (From http://rs.tdwg.org/dwc/)
Darwin Core Archive (DwC-A) | A file archive (.zip) that contains structured *metadata* and primary data in the form of *two-dimensional data tables* combined into a *relational database* via *identifier* columns. 
Occurrence | 
Measurement | 
Aggregated Value |
Value | 
Fact | 
Method handbook | 
File hosting service | An online platform that hosts *datasets* or entire *repositories* and provides access to a wide audience on the internet; popular examples in biology are Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org;

Table: Glossary of terms from the biodiversity data-management context as they are used in this paper. (draws from @garnier16)

\normalsize

# A review of initiatives for trait-data standardisation

In this section, we review four types of initiatives that are of relevance for trait data standardisation:

1. Initiatives that provide *trait datasets* which have been assembled out of a particular research interest, either by measurement or collated from the literature. 
2. Initiatives that aim to harmonise trait data from the literature or from direct measurement into *trait databases* and make those data widely available. 
3. Initiatives that aim at the standardisation and development of consensus measurement methods and definitions for traits, and provides standard *terminologies* in the form of *thesauri* and *ontologies*.
4. Initiatives that aim to leverage *relational database* structures and *semantic web* technology to link trait data to a wider set of biodiversity data.

We discuss these initiatives separately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17]. 
We show how the degree of trait-data standardisation in existing datasets spans this entire spectrum and which tools and standards are applied to achieve harmonisation of data from multiple, distributed sources. 
The objective of this review is to raise awareness for the generic structure of trait data and aid researchers to share and publish own datasets in an appropriate form.

## Trait datasets

In the field of comparative biology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been published in regular journal articles or books. 
With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @gossner15; @ricklefs17].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->

Today, the online publication of such data is greatly facilitated by file hosting services (e.g. Figshare.com). 
Furthermore, these services warrant long-term accessibility, offer DOIs for referencing, and support Public Domain dedication or Creative Commons licenses.
However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk.

Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds for *metadata* documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources for data management. 
As a result, datasets greatly vary in structure and terminology.

For trait-data, there are typical issues arising from the variability of data structures. 
For instance, the column descriptions andterminology applied to taxa and traits are project specific, and rarely chosen to allow translation into larger database initiatives. Furthermore, metadata varies in its detail, e.g. for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
In terms of structure, trait data usually are reported in a species$\times$trait wide-table format. In this format, each row contains a species (or taxon) for which multiple traits are reported in columns. 
Similarly, when reporting raw data, researchers place observations of individual organisms in rows with multiple trait measurements applied to the same individual across multiple columns. 
Variability in the number and meaning of columns in these *data tables* requires tedious manual adjustments when merging multiple datasets [@wickham14].

<!--To remedy these problems, data management initiatives that arise from scientific communities and projects, like the KNB (the Knowledge Network for Biocomplexity; https://knb.ecoinformatics.org/) or GFBio [the German Federation on for Biological Data; gfbio.org; @diepenbroek14], offer means for small projects to meet the high scientific criteria of data management.-->

A global overview of existing trait data for all taxa and trait types is difficult to obtain. Therefore, in an attempt to collate a list of existing, distributed and unstandardised datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use (see Appendix A for a current excerpt of this list). We invite data owners and users to add further trait datasets to this spreadsheet.

As it stands, the decentralisation and the lack of data standardisation of low-threshold online repositories renders the compilation of data into larger collections inefficient and reduces the potential of many published datasets to be re-used and combined into broad synthesis analysis.

## Database initiatives

<!-- community database backend, but static publication --> 
In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g. @klotz02; @iversen17; @kleyer08; @oliveira17; @myhrvold15; @kissling14; @jones09].<!--[@wright04; ] -->
While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets apply a standardised terminology for taxa or traits that would allow them to be efficiently related to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary. -->
Documentation of metadata and methodology differs in the level of detail, depending on the research focus of the initiative. 
Just as the individual datasets described above, many of these databases are published as static data tables on low-threshold file hosting platforms and are updated irregularly.

As they deal with much larger amounts of data, initiatives that form around natural history museum collections are more concerned with data standardisation. 
With the digitization efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowdsourcing (e.g. www.markmybird.org), data on body measurements are likely to grow exponentially in the near future. 
For example, the VertNet database compiled and harmonized large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources are integrated.

<!-- database backend, selective download, user access management --> 
More specialised trait-database platforms have been created to cover certain trait types (e.g. floral traits, seed traits, root traits or wood density traits), interaction types (e.g. pollination traits or feeding relationships), or a specific environmental and experimental context of the trait observation (e.g. location or climatic data).  
Such database initiatives attract data submissions from a defined research field and take care of the harmonisation process and thereby greatly facilitate data synthesis.
<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
For example, by aiming for a universal framework for plant traits, the TRY database for plant traits [@kattge11]  attracted more data submissions and downloads than any other trait data platform. The *online database* enables selective data download and user permission and rights management. As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section). 
Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12]. Here, taxa are derived from metagenomic analysis [@torsvik02; @langille13]. Databases are also used to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g. the KEGG orthology, @kanehisa12]. 
For animals, a single unified platform and harmonizing scheme for animal trait data is still lacking. The reason for this may be that harmonising trait data on animals, which span multiple trophic levels and possess diverse body plans, is a more complex task than for plants [@moretti17]. Nonetheless, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).

Regarding open access, few of these centralised databases comply with the criteria demanded by journals and funding agencies for primary data publication. The platforms incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and potentially co-authorship.
With the proactive turn towards open access data (as stated in the Bouchot Declaration; http://www.bouchoutdeclaration.org/), it may be necessary to find other incentives for data submission.

<!-- AO: In a sense of data availability to the public : yes. But do we discuss it here? We want to show how to make trait data sharing easier due to better data format, or?Not the topic of this article to discuss open vs. non-open data, or?
bKR: I am not quite sure what point you are trying to make it, the paragraph sort of goes back and forth between advantages and drawbacks of the different approaches. 
FDS: I think its part of the rationale for a simple data standard. More open data mean that we need more awareness for data standardisation. 
 -->

## Thesauri and Ontologies for traits

A major challenge in trait-data standardisation is the lack of widely accepted and unambiguous trait definitions. Previous standard definitions of trait *concepts* range from listings of selected definitions in *glossaries*, over well-defined methodological *handbooks* and and comprehensive *thesauri*, to relational definitions of trait concepts in *ontologies*. 
While glossaries may be seen as specific for a study context, the initiatives behind method handbooks, thesauri and ontologies are primarily concerned with consensus building on trait definitions in a wider community.

Assigning an unambiguous methodological protocol to a trait is key for standardising the physical process of measuring. Efforts to develop handbooks for measurement protocols provide such a methodological standardisation for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but obviously are of limited use in harmonising trait data that predate or ignore this standard [@kattge11a].
Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].

A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
Expanding on this, ontologies link the defined terms to each other by formally defining the relationships between them, with the objective of enabling a computational interpretation of data. 
Being publicly available, it is also possible to refer to these defined terms via globally unique *Uniform Resource Identifiers (URIs)* within own datasets. For example, a measurement of seed size could be linked to the Planteome Trait Ontology definition of 'seed size' by referencing 'http://browser.planteome.org/amigo/term/TO:0000391'. 
Ontologies define terms based on other well-defined terms from published ontologies. The Planteome Trait Ontology definition of the concept 'seed size' contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
This way, each trait definition may link to a broader or narrower term. For example, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'. Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies. 
By providing this interlinkage of trait ontologies, a web of definitions is spun across the Internet which allows researchers and search engines to relate independent trait measurements with each other and connect it to the wider *semantic web* of online data [@berners-lee01; @page08]. 
The distinction of thesauri and ontologies is not truely binary. Rather they mark idealised ends of a spectrum. While thesauri may contain defined relations between terms within the standard, ontologies relate most terms to other defined concepts, and also link those to other standards.

Comprehensive trait thesauri have been developed in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], and in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14]. 
Ontologies of trait definitions have been developed for plants [e.g. the Plant Ontology, @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g. the Hymenoptera Anatomy Ontology, @yoder10; the vertebrate trait ontology, @park13]. 
The existing thesauri and ontologies for traits differ widely in terms of hierarchical depth and detail, as well as in curation efforts and measures for peer-reviewed quality control.

<!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g. units or legit factor levels) of measured values or reported facts. It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits. --><!-- BKR: Isn’t that the same? Or can you have multiple parents for one node? FDS: agree. simplify phrasing. -->

Meta-ontology initiatives, like Planteome.org, offer access to multiple published ontologies and build platforms for the collaborative development [@walls12]. 
For general biodiversity data, the OBO Foundry (http://www.obofoundry.org/), ontobee (http://www.ontobee.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer webservices for computational access.  <!-- Mention also NCBO BioPortal? https://bioportal.bioontology.org/ -->

To conclude, there is already a suite of globally available thesauri and ontologies for traits that emerged from standardisation efforts of methodologies and community consensus processes. Albeit definitions in some domains are better covered than others. Interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards.
Most importantly, while these defined vocabularies are widely used in biodiversity data management, distributed data repositories of smaller project contexts hardly make use of them. 
A more widespread implementation of ontologies would advance the possibilities to aggregate datasets into databases and reduce noise and uncertainty. 
To achieve this, the use of ontologies and thesauri must be incentivised and facilitated for individual researchers.
For example the accessibility of ontologies will increase if advanced webservices, and particularly open Application Programming Interfaces (APIs), are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools. 
Software then can assist researchers in linking own data to globally defined concepts.

## Trait-data structures for the semantic web

While trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, e.g. for taxa. 
To make sense of trait data in the context of more general databases, a consensus definition of trait data is necessary.

Trait data have been defined by @garnier17 to follow an entity-quality model (EQ), where a trait observation is 'an entity having a quality'. 
More specifically, a trait dataset contains information on quantitative *measurements* or qualitative *facts* (i.e. trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or parts of an individual specimen) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
We are expanding on this definition: quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig. 1*a*), by aggregating replicated measurements on multiple entities (Fig. 1*b*) or by estimating the means or ranges for the respective taxon as reported in the literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
The entity or observation (i.e. the *occurrence*) to which the reported measurement or fact applies may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g. a genus).

\begin{figure}[t]
\centering
\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
\caption{Types of ecological trait data assume different entities or reported quantities. a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g. reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g. a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire taxon.}

\end{figure}
<!-- DF : The “variables” for the taxa etc should only be reused for the different examples when they refer to something similar: “a) Entity: an individual X of taxon Y”, “b) Entity: taxon Y”, “c) Entity: males of taxon Z” and “d) Entity: genus U”. Alternatively, since these are already example, why use variables here in the first place and not just specific example for the taxa as well: “a) Entity: an individual X of taxon Tettigonia viridissima” -->

These relationships between a trait observation and an individual organism as an occurrence of a particular taxon have been formalised in the schema for biological collection records [ABCD Schema; @walls14] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12]. For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) applies these terms.
These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining names for columns that contains measurement values, units, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details. The terminologies provided by these standards are quite universal and even cover most use cases of trait data. An entire ecosystem of data standards links to and expands the capacities of DwC [@wieczorek12].  
<!--  DF: am not sure, where to best write this comment, but I think it belongs somewhere in this chapter. The Trait Data Standard is a standard on how to store trait data measurements. In this chapter you mention what databases for trait data there are (usually specific for certain taxa) and what ontologies there for the different traits, however it is missing the section on how trait data measurements have been and can be stored prior to your standard. For instance the ABCD standard has a huge section for MeasurementsAndFacts. And while it certainly is not primarily made to capture trait measurements, it can certainly be used for it (at least on a unit level, combining measurements for higher taxa is not possible). Darwin Core has similar, though not as extensive, fields for measurements, as they were modelled based on ABCD. While I agree that your standard is useful and necessary, it certainly is not the first standardized way to describe trait measurements.  -->

Specifically designed for plant traits, @kattge11 proposed a generic database structure that covers most potential use cases of trait-based ecology. This data structure is built around a central data table that contains observations, i.e. a single event of measurement on the same individual plant specimen at the same point in time. This structure emphasises the fact that multiple trait data are measured on the same individual organisms and used to analyse correlations between these multiple traits. Identifiers link the measurements (qualities) to the same observation (entity), each measurement being well defined by additional standard tables. The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment. This structure can be implemented in any relational database management system.

In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms of life, which is to date the most general approach to providing an integrated structure for trait data.
The framework employs established terms provided by the DwC, relates trait definitions to trait ontologies for phenotypic or anatomical terms, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives]. 
Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09a], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/]. 
The archive also integrates the general framework for metadata of the Ecological Metadata Language [EML, @knb11]. 
The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review [@franz16]. Initiatives that aim at providing a stable reference for taxa are for instance the EOL Catalogue of Life [http://www.catalogueoflife.org/, @roskov18], the GBIF Backbone Taxonomy [@gbifsecretariat17], or the EDIT Platform for Cybertaxonomy (https://cybertaxonomy.eu/).

>>>>

---
title: "Towards an Ecological Trait-data Standard - Appendix A"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2] \footnotemark[3]
- name: Malte Jochum \footnotemark[4]
- name: Gaetane Le Provost \footnotemark[3]
- name: Andreas Ostrowski \footnotemark[5]
- name: Caterina Penone \footnotemark[4]
- name: David Fichtmüller \footnotemark[5]
- name: Anton Güntsch \footnotemark[5]
- name: Pete Manning \footnotemark[3]
- name: Birgitta König-Ries\footnotemark[6]
- name: Nadja K. Simons \footnotemark[7]
fontsize: 12pt
geometry: margin=0.5in
classoption: landscape
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
---

@@

\newpage

```{r, echo = FALSE, warning = FALSE}
options(knitr.table.format = "latex")
library(kableExtra)
gg <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTfQOIHjJv5x_YHgq14DR7VLTvycfsl2Njj2Ip4p2a5kca8AX5t5cmaial6Bg-MX9pBDOJaJfnlR7Qj/pub?gid=0&single=true&output=csv", encoding = "UTF-8")
dt <- subset(gg, select = c(datasetName, type, DOI, NumberTaxa, NumberTraits, Responsible, CiteAs, TermsOfUse))

kable(dt, longtable=TRUE, 
      caption = "List of published trait datasets and databases", 
      booktabs = T) %>%
kable_styling(latex_options = c("scale_down", "repeat_header"), font_size = 7) %>%
  column_spec(1, width = "15em") %>%
  column_spec(2, width = "5em") %>%
  column_spec(3, width = "10em") %>%
  column_spec(6, width = "10em") %>%
  column_spec(7, width = "25em") %>%
  column_spec(8, width = "11em") 
```
